Using Dependent Types to Build Safe and Concise JavaScript Foreign Function Interfaces

### 2013-06-07 papers/sle2013/sle2013.tex
sle code examples

\section{Motivating Example}

Typical tasks involved in Web applications.

@@

>>>>

\section{Background Material}

@@

\subsection{DOM Creation}

\begin{lstlisting}[language=JavaScript]
var div = document.createElement('div');
var input = document.createElement('input');
\end{lstlisting}

\code{div} has type \code{DivElement} while \code{input} has type \code{InputElement}. These types have different
methods. As stated in the introduction, most statically typed language have an API that returns an \code{Element},
which is the least upper bound of the possible types returned by \code{createElement}, forcing users to explicitly
down-cast the result to the expected type, thus loosing type safety.

Alternatively, some languages use overloading.

Consider the following function creating an element and giving it a class name:

\begin{lstlisting}[language=JavaScript]
var create = function (elementType, className) {
  val el = document.createElement(elementType);
  el.className = className;
  return el
};
create('input', 'field').focus();
create('img', 'figure').src = 'http://google.com/logo.png';
\end{lstlisting}

How to type check the above code? The problem is that the return type of the \code{create} function depends on its
first parameter.

Note that a possible solution in Java could be the following:

\begin{lstlisting}[language=Java]
class ElementName<E> {}
ElementName<InputElement> Input = new ElementName<InputElement>();
ElementName<ImageElement> Img = new ElementName<ImageElement>();

<E> E create(ElementName<E> type, String className) {
  ...
}

create(Input, "field").focus();
create(Img, "figure").setSrc("http://google.com/logo.png");
\end{lstlisting}

However, the type parameter \code{E} has to be filled at usage site. If the compiler can infer it, that’s fine.

\subsection{DOM Events}

A similar issue applies to the DOM events API.

\begin{lstlisting}[language=Java]
class EventName<E> {}
EventName<ClickEvent> Click = new EventName<ClickEvent>();

<E> Stream<E> on(EventName<E> type, Callback<E>) {
  ...
}

Callback<ClickEvent> incrementCounter = new Callback<ClickEvent>() {
  @Override
  void execute(ClickEvent event) {
    ...
  }
}

on(Click, incrementCounter);
\end{lstlisting}

Again, the type parameter \code{E} has to be filled at usage site.

The problem with selectors is slightly different.

\begin{lstlisting}[language=JavaScript]
var input = document.querySelector('input');
var el = document.querySelector('#content');
\end{lstlisting}

\code{input} can not have another type than \code{InputElement}, but \code{el} can be an element of any type (depending
on which element has the id \code{content}). Return type can not \emph{always} be inferred from the parameters.
In this case, down-casting is impossible to avoid, but how to make it optional, and how to make it less risky?

We want to be able to write the following code:

\begin{lstlisting}
val input = document.find(Input) // input has type InputElement
val el = document.querySelector("#content") // el has type Element
val img = document.querySelector[Img](".figure") // note that the type Img is explicitly applied in this case but was automatically inferred in previous cases
val error = document.querySelector[Int]("div") // Does not compile: querySelector can not return a number
\end{lstlisting}
